A Formal Verification Framework for Tezos Smart Contracts Based on Symbolic Execution

### 2024-08-23 papers/static-checker/aplas/paper.tex
cosmetics & spell checking

%
\authorrunning{Ha Doan, P. Thiemann}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Freiburg, Germany \\
  \email{\{doanha,thiemann\}@informatik.uni-freiburg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We developed SCV, a verification tool geared towards contracts implemented in Michelson, the smart contract language of the Tezos blockchain. SCV utilizes symbolic execution to derive verification conditions that can be checked automatically against user specifications using an SMT solver. These specifications are written in a domain-specific language. We present the fundamental principles behind the design and development of SCV. By pruning states during symbolic execution, SCV effectively addresses the problem of state explosion. This approach has enabled us to conduct two case studies on real-life contracts, demonstrating the expressiveness of our tool.
\keywords{Smart Contract, Blockchain, Formal Verification, Symbolic Execution}
\end{abstract}
\section{Introduction}
\label{sec:introduction}
Since smart contracts are permanently stored on the blockchain, once deployed, a smart contract cannot be changed. Regrettably, numerous instances have been documented where vulnerabilities in smart contracts were identified by the blockchain community, resulting in substantial financial losses~\cite{dao,wallethack}. The distinctive execution model of smart contract languages has given rise to unforeseen errors due to a lack of familiarity with their complexities. These factors, combined with the immutability of contracts, underscore the need to ensure their correctness prior to deployment.

We have developed SCV, a formal verification tool for smart contracts on the Tezos blockchain \cite{tezos-whitepaper}, written in Michelson \cite{michelson}. SCV relies on symbolic execution backed by an SMT solver to simulate the implementation of the smart contract language, helping to detect subtle errors. Additionally, SCV includes a domain-specific language for precisely specifying smart contract properties, tailored to Michelson's execution model and inspired by Z3. Each specification consists of Michelson code, a precondition, and a postcondition with the usual partial correctness semantics. Our symbolic interpreter calculates a postcondition predicate from the specified precondition. The SMT solver Z3 \cite{z3} then checks that this predicate implies the specified postcondition.

While there is already a considerable body of work applying symbolic execution to the verification of smart contracts \cite{securify,manticore,kevm,park} (see Section~\ref{sec:related-work} for details), most of these efforts are designed for Ethereum, where standard techniques for symbolic execution of imperative programs are applicable. Such groundwork does not exist for Michelson, which is a functional language running on a stack machine (see Section~\ref{sec:background}) with a sophisticated instruction set that includes high-level loops. For that reason, we had to develop symbolic execution for Michelson from the ground up.

Besides the Mi-Cho-Coq framework that relies on interactive theorem proving \cite{micho}, HELMHOLTZ \cite{helmholtz} is another automatic verifier for Michelson. However, the approach of HELMHOLTZ is significantly different from ours. HELMHOLTZ defines a refinement type system to keep track of the properties of values in the storage and on the stack, so that verifying a program corresponds to type-checking against a specification type. Furthermore, these tools generally conduct experiments with sample smart contract codes containing about 100 instructions. The HELMHOLTZ experiment primarily replicates issues previously identified by Mi-Cho-Coq. Although HELMHOLTZ analyzes one real-world contract, it only examines a portion of the code. In contrast, our tool is capable of handling entire contracts with hundreds or even thousands of instructions.

SCV utilizes an SMT solver to eliminate invalid states during symbolic execution, addressing the problem of state explosion. This capability enabled SCV to conduct two case studies involving real smart contracts that have been operational on the Tezos blockchain for quite some time. Despite extensive testing, a flaw was reported in the Kolibri oracle contract. In conclusion, SCV streamlines the process of uncovering hidden errors and validating user-defined properties. It provides a comprehensive solution to mitigate potential pitfalls in blockchain-based applications.

Our contributions are as follows.
\begin{enumerate}
\item Design and implementation of the SCV domain specification language.
\item Design and implementation of a symbolic interpreter for a significant subset of Michelson. 
\item Two case studies verifying smart contracts on the Tezos blockchain:
  \begin{itemize}
  \item USDtz, an implementation of the FA~1.2 standard for financial applications;
  \item Kolibri, an oracle contract that injects currency exchange rates from an external system.
  \end{itemize}
\end{enumerate}

Several high-level languages, such as LIGO \cite{ligo}, SmartPy \cite{smartpy}, and Liquidity \cite{liquidity}, provide a more readable syntax resembling popular programming languages. This allows developers to write Tezos smart contracts more comfortably. Our tools are applicable to all these languages because they compile to Michelson.
\section{Michelson}
\label{sec:background}
Michelson is a stack-based, statically typed functional programming language. A program is defined by the type $S$ of its storage, the type $P$ of its single parameter, and a typed list of instructions for the underlying stack machine. It starts on a stack with one element containing a pair of the current storage value (taken from the blockchain) and the parameter value specified by the contract invocation. When the program finishes, the stack must contain a single element with a pair of the final storage value (to be saved on the blockchain) and a list of operations. These operations are syntactic commands, such as token transfers or contract invocations, among other possibilities. Thus, the program computes a function $P \times S \to O^* \times S$, where $O$ is the type of operations.

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions) with the expected operations, Michelson supports a number of types and operations specific to blockchain applications (e.g., \texttt{mutez} for tokens) as shown in Fig.~\ref{fig:types}, where \textbf{cpt} denotes comparable types. The type system does not support quantification over types, and there are no user-defined types.
\begin{figure}[tp]
\begin{align*}
\textbf{cpt} &::= 
   \text{unit} 
   \Mid\text{never} 
   \Mid\text{bool} 
   \Mid\ \text{int}
   \Mid\text{nat}
   \Mid\text{string}
   \Mid\text{chain-id} 
    \Mid\text{bytes}
   \Mid\text{mutez} 
   \Mid\text{key-hash} \\
   &\Mid\text{key}
   \Mid \text{signature}
   \Mid \text{timestamp} 
   \Mid \text{address} 
   \Mid \text{option}\ \textbf{cpt}
   \Mid \text{or}\ \textbf{cpt}\ \textbf{cpt}
   \Mid \text{pair}\ \textbf{cpt}\ \textbf{cpt} \\
\text{T, U} &::= 
    \text{cpt}
   \Mid \text{option}\ \text{T}
   \Mid \text{list}\ \text{T}
   \Mid \text{set}\ \textbf{cpt} 
   \Mid \text{operation} 
   \Mid \text{contract}\ \text{T} \\
  & \Mid \text{pair}\ \text{T}\ \text{U}
   \Mid \text{or}\ \text{T}\ \text{U}
   \Mid \text{lambda}\ \text{T}\ \text{U} 
   \Mid \text{map}\ \textbf{cpt}\ \text{T}
   \Mid \text{big-map}\ \textbf{cpt}\ \text{T}
\end{align*}
\caption{Types}
\label{fig:types}
\end{figure}

Listing~\ref{lst:simple-program-add} contains a short example of a Michelson program. The code is interspersed with comments indicating the types of values currently on the stack. This program first duplicates its input (\lstinline|DUP|), decomposes the pair $P \times S$ into its components (\lstinline|UNPAIR|), compares the two values on top of the stack (\lstinline|COMPARE|), and transforms the result into a boolean (\lstinline|GE|). The \lstinline|IF| instruction pops the boolean and either executes the first (for true) or second (for false) list of instructions. At this point, only the initial pair remains on the stack. In the true branch, \lstinline|UNPAIR| decomposes the pair, and \lstinline|SUB| subtracts the two components. In the false branch, the \lstinline|PUSH| instruction puts a string message on the stack. The \lstinline|FAILWITH| instruction aborts the contract execution: the tokens provided with the invocation are consumed, but the storage and balance of the contract remain unchanged. Finally, the \lstinline|NIL| instruction pushes an empty list of operations, and \lstinline|PAIR| creates the required pair of type $O^* \times S$.

A contract may have several entrypoints. Entrypoints are encoded by
using a sum type (call ``or type'' in Michelson). Each (nested) alternative in a sum can be
given a name, which is used as the name for an entrypoint. A call to
an entrypoint automatically adds the required \lstinline|LEFT| and
\lstinline|RIGHT| constructors to the parameter.
\lstset{language=michelson}
\begin{lstlisting}[float,captionpos=b,caption={A Michelson program},label={lst:simple-program-add}]
parameter int;
storage   int;
code {                (* pair int int :: [] *)
  DUP; UNPAIR;        (* int :: int :: pair int int :: [] *) 
  COMPARE; GE;        (* bool :: pair int int :: [] *)
  IF {UNPAIR; SUB} {PUSH string 'Unexpected Pair'; FAILWITH};
                      (* int :: [] *)
  NIL operation; PAIR (* pair (list operation) int :: [] *)}
\end{lstlisting}
\section{SCV: A Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}
SCV contains a domain-specific language to specify smart contract properties, aiming to be user-friendly for non-experts and compatible with SMT solvers. 
\subsection{Basic syntax}
\label{sec:basic-syntax}
Listing~\ref{lst:sub-contract-specification} shows the specification of the \texttt{sub} contract in the SCV language. Here, \KMCONTRACT, \KSPEC, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KINPUT{} and \KOUTPUT{} fields specify the input and output patterns, which are terms on Michelson values.

Figure~\ref{fig:syntax-terms-predicates} contains the syntax of terms: A term \textbf{t} can consist of constants of the sort \text{nat} (natural number), \text{int} (integer), \text{string} (string), and \text{b} (bytes) along with their respective types. The type annotation may be omitted except for the type \text{nat} to distinguish it from integers. It also encompasses all blockchain constants \textbf{c}. When a variable appears for the first time, it must be accompanied by its type declaration. Subsequent mentions can omit the type declaration.

Since Michelson operates on a stack, we assume that the input stack and the output stack each have only one element, as specified in the input term and the output term, respectively. The syntax allows us to specify either a fragment of Michelson code or a complete program. For a complete Michelson program, the output should be a pair consisting of an operation list and the state of the storage after code execution,  as described in Section \ref{sec:background}. For verification purposes, we elide the operation list and represent the output solely by the term of the value in the storage after execution concludes.

The pre and postcondition instances are formulas in first-order logic that can be handled by an SMT solver (specifically, Z3 in our case). The specification language is constructed from predicates. Formulas used in assertions or added to solvers are terms of Boolean type. Non-Boolean terms may also be used, and they can be mixed in any combination where types match appropriately.

Several predefined functions include \text{get\_map(e1, e2)}, which accesses an element in a map. If there is no key \text{e1} in the map \text{e2}, the function returns \text{None}; otherwise, it returns \text{Some(v)} where $v$ is the mapped value. Another function, \text{get\_contract(T, e)}, converts the address \text{e} to a contract on the Tezos blockchain. If no contract with an entrypoint of type \text{T} is associated with the address \text{e}, it returns \text{None}; otherwise, it returns the contract as \text{Some(c)}.

Certain functions are defined as predicates, such as \text{transfer\_token(e1, e2, e3)}. This predicate holds true if, after executing the smart contract code, there is an operation in the operation list indicating the transfer of \text{e2} tokens to address \text{e3} along with parameter \text{e1}\footnote{Additional predicates can be straightforwardly added if a mapping to Z3 formulas is defined.}.
%\lstset{language=michelson}
\begin{lstlisting}[float,captionpos=b,caption={Specification of  the \lstinline/sub/ contract},label={lst:sub-contract-specification},numbers=left]
mcontract sub = spec
  code := {DUP; UNPAIR; COMPARE; GE; IF {UNPAIR; SUB} 
          {PUSH string 'Unexpected Pair'; FAILWITH}}
  input := Pair (x : int) (y : int)
  output := (z : int)
  pre-condition := (x >= y)
  post-condition := (z = x - y) && (z >= 0)
\end{lstlisting}

In Listing \ref{lst:sub-contract-specification}, the program takes as input a stack with only one

@@

The specification continues by defining the output term as a variable \lstinline/z/ of interger, and 
pre and postconditions, where the precondition states the assumption
that the integer \lstinline/x/ is greater than or equal to  \lstinline/y/. The postcondition asserts that the result

@@

Let us return to the auction contract. It is important to guarantee
that  the bid call can happen successfully after another if the new bid is higher than the previous
one (lines 16-17). The  \lstinline/%close/ entrypoint can be called
anytime after bidding if the caller is the contract owner (lines 18-19). However, it
should not be possible to call the  \lstinline/%bid/ entrypoint after
the auction is closed, and, of course, it cannot be closed
again if it is already closed (line 20).  Furthermore, contract deployment sets the  \lstinline/open/ flag to
\TRUE, indicating that the contract is open for bidding, i.e., users
should be able to bid on the contract or the contract owner should be
able to close it. Otherwise, the owner's original
deposit is lost. This sequence  illustrates
the life cycle of the auction contract, which is depicted in Figure
\ref{fig:auction-life-cycle}. Other properties related to the \lstinline|%close| entrypoint and the life cycle are omitted here due to page limits.
\begin{figure}[tp]
    \centering
    \includegraphics[width=0.9\textwidth]{auction}
    \caption{Auction contract life cycle}
    \label{fig:auction-life-cycle}
\end{figure}

@@

A state is unreachable if its predicate \PREDICATE\ is  not
satisfiable. The following rule drops unreachable states.
\begin{mathpar}
\inferrule[]
  { \UNSAT\ \PREDICATE
  }{
  \{[\INSTRUCTION, \STACK, \PREDICATE]\} \cup \SYSTEM \SystemTrans \SYSTEM}
\end{mathpar}
The symbolic interpreter works as a graph with states as nodes and transitions as edges. This representation simplifies performing multi-step subtransitions needed for some operations.

The semantics of the symbolic rules for the instructions adhere to the formal definitions of the Michelson language as outlined in \cite{michelson,michelson1}. In our model, these instructions are classified based on how we handle them into one-step, multi-step, blockchain, cryptographic, branch, and loop instructions.

@@

Branch instructions, such as \IF\{\INSTRUCTIONONE\}\{\INSTRUCTIONTWO\} consume a value at the top of the stack. If the branch condition
is True, it executes \INSTRUCTIONONE; otherwise, it executes
\INSTRUCTIONTWO. 
\begin{mathpar}
  \inferrule[]
  { }{
    [(\IF\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TBOOL) \STACKCONCAT\STACK, \PREDICATE]
    \StateTrans\
    \{[\INSTRUCTIONONE, \STACK, \PREDICATE\ \Wedge\ \StackOne],   [\INSTRUCTIONTWO, \STACK, \PREDICATE\ \Wedge\ \NEG\
   \StackOne]\}} 
\end{mathpar}

@@

Let us first take look at the terminate loops, such as the \ITER\ instruction, which iterates over a
list. The corresponding typing rule for the \ITER\ instruction is as follows: 
\begin{mathpar}
  \inferrule{\JTypeExpr\TEnv{\INSTRUCTION}{\TY \STACKCONCAT \TYA\ \SRightarrow\ \TYA}
  }{
      \JTypeExpr\TEnv{\ITER\ \INSTRUCTION}{\TYLIST\ \TY \STACKCONCAT \TYA\ \SRightarrow\ \TYA}
    }
\end{mathpar}
This rule ensures that if the loop body \INSTRUCTION\ has the type \TY\ \STACKCONCAT\ \TYA\ \SRightarrow\ \TYA, then the entire \ITER\ instruction, when applied to a stack with the top element of type \TY\ \TYLIST\ and the rest of the stack of type \TYA, produces a result stack of type \TYA.

@@

In separate work \cite{dynamic-logic}, we have designed a dynamic logic tailored to Michelson as a formal foundation for the implementation of SCV. We have formalized a soundness proof of this logic in the Agda proof assistant.
\section{Static Checker}
\label{sec:static-checker-we}
The architecture of the static checker is illustrated in Figure~\ref{fig:architecture-of-static-checker}. Users write specifications in the SCV language, which are then parsed into an Abstract Syntax Tree (AST) in OCaml. The input term along with the code is passed to the symbolic interpreter, while the output term and the pre and postconditions are fed to the static checker. The symbolic interpreter, which is fully typed and rejects ill-typed programs, then runs symbolic execution on the code with the provided input, where the initial stack is derived from the input term.

@@

We have implemented approximately 85\% of the Michelson instructions, including most of the major-use instructions. The remaining instructions were not implemented primarily due to time constraints rather than technical limitations.

Consider the specification of the \lstinline/sub/ contract, the results of
the symbolic execution are either a failwith state with the path
condition \lstinline/x/ $<$ \lstinline/y/,  or a state in which the
final stack has only one element, which is a symbolic variable
\lstinline/sbvar/ of type integer, with the path condition stating
that the variable \lstinline/x/ is greater than or equal to
\lstinline/y/ and  \lstinline/sbvar/ is equal to
the subtract of \lstinline/x/ and \lstinline/y/. 
\subsection{Output check}
\label{sec:output-check}
For a fragment of Michelson code, the output term of the symbolic execution has the form of a term \Term\ and its type \TY. For a complete smart contract, it takes the form \PAIR\ \VOPERATIONLIST\ \VSTORAGE, \TPAIR\ (\TOPERATIONLIST) \TYS, where \VSTORAGE\ represents the storage and \VOPERATIONLIST\ represents the operation list. Given an output term specified as $\tau$, the checker scans all reachable final states that do not contain a failwith stack and checks whether $\tau$ matches with the output term. In the former case, it matches $\tau$ with the term \Term. In the latter case, it matches with \VSTORAGE.

>>>>

%
\authorrunning{Ha Doan, P. Thiemann}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Freiburg, Germany \\
  \email{\{doanha,thiemann\}@informatik.uni-freiburg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We developed SCV, a verification tool geared towards contracts implemented in Michelson, the smart contract language of the Tezos blockchain. SCV utilizes symbolic execution to derive verification conditions that can be checked automatically against user specifications using an SMT solver. These specifications are written in a domain-specific language. We present the fundamental principles behind the design and development of SCV. By pruning states during symbolic execution, SCV effectively addresses the problem of state explosion. This approach has enabled us to conduct two case studies on real-life contracts, demonstrating the expressiveness of our tool.
\keywords{Smart Contract, Blockchain, Formal Verification, Symbolic Execution}
\end{abstract}
\section{Introduction}
\label{sec:introduction}

Once a smart contract is deployed on the blockchain, it becomes permanent and cannot be altered.
Regrettably, numerous cases have been documented where vulnerabilities
in smart contracts were identified and exploited by bad actors in the blockchain community, resulting in substantial financial losses~\cite{dao,wallethack}. The distinctive execution model of smart contract languages has given rise to unforeseen errors due to a lack of familiarity with their complexities. These factors, combined with the immutability of contracts, underscore the need to ensure their correctness prior to deployment.

We have developed SCV, a formal verification tool for smart contracts
on the Tezos blockchain \cite{tezos-whitepaper}, written in Michelson
\cite{michelson}. SCV relies on symbolic execution backed by an SMT
solver to simulate the implementation of the smart contract language,
helping to detect subtle errors. Additionally, SCV includes a
domain-specific language for precisely specifying smart contract
properties, tailored to Michelson's execution model and informed by
the capabilities of Z3. Each specification consists of Michelson code, a precondition, and a postcondition with the usual partial correctness semantics. Our symbolic interpreter calculates a postcondition predicate from the specified precondition. The SMT solver Z3 \cite{z3} then checks that this predicate implies the specified postcondition.

While there is already a considerable body of work applying symbolic
execution to the verification of smart contracts
\cite{securify,manticore,kevm,park} (see
Section~\ref{sec:related-work} for details), most of these efforts are
targeted at Ethereum, where standard techniques for symbolic execution
of imperative programs are applicable. Such groundwork does not exist
for Michelson, which is a functional language running on a stack
machine (see Section~\ref{sec:background}) with a sophisticated
instruction set that includes high-level loops. For that reason, we
had to develop symbolic execution for Michelson from scratch.

Besides the Mi-Cho-Coq framework that relies on interactive theorem
proving \cite{micho}, HELMHOLTZ \cite{helmholtz} is another automatic
verifier for Michelson. However, the approach of HELMHOLTZ is
significantly different from ours. HELMHOLTZ defines a refinement type
system to keep track of the properties of values in the storage and on
the stack, so that verifying a program corresponds to type-checking
against a specification type.
Furthermore, the reported experiments with these tools consider sample
smart contract codes of about 100 instructions. The HELMHOLTZ
experiments primarily replicate issues previously identified by
Mi-Cho-Coq. Although HELMHOLTZ analyzes one real-world contract, it
only examines a portion of the code. In contrast, our tool is capable
of handling entire contracts with hundreds or even thousands of
instructions.

SCV utilizes an SMT solver to eliminate invalid states during symbolic
execution, addressing the problem of state explosion. This capability
enabled us to conduct two case studies involving real-world smart contracts
that have been operational on the Tezos blockchain for quite some
time. Despite extensive testing, a flaw was reported in the Kolibri
oracle contract. In conclusion, SCV streamlines the process of
uncovering hidden errors and validating user-defined properties. It
provides a comprehensive solution to mitigate potential pitfalls in
blockchain-based applications.

Our contributions are as follows.
\begin{enumerate}
\item Design and implementation of the SCV specification language.
\item Design and implementation of a symbolic interpreter for a significant subset of Michelson. 
\item Two case studies verifying smart contracts deployed on the Tezos blockchain:
  \begin{itemize}
  \item USDtz, an implementation of the FA~1.2 standard for financial applications;
  \item Kolibri, an oracle contract that injects currency exchange rates from an external system.
  \end{itemize}
\end{enumerate}

Several high-level languages for smart contracts, such as LIGO
\cite{ligo}, SmartPy \cite{smartpy}, and Liquidity \cite{liquidity},
support writing smart contracts for Tezos. Our tools
are applicable to all these languages because they compile to
Michelson. 
\section{Michelson}
\label{sec:background}
Michelson is a stack-based, statically typed functional programming language. A program is defined by the type $S$ of its storage, the type $P$ of its single parameter, and a typed list of instructions for the underlying stack machine. It starts on a stack with one element containing a pair of the current storage value (taken from the blockchain) and the parameter value specified by the contract invocation. When the program finishes, the stack must contain a single element with a pair of the final storage value (to be saved on the blockchain) and a list of operations. These operations are syntactic commands, such as token transfers or contract invocations, among other possibilities. Thus, the program computes a function $P \times S \to O^* \times S$, where $O$ is the type of operations.

Besides the standard datatypes (e.g., numbers, lists, pairs, sums, functions) with the expected operations, Michelson supports a number of types and operations specific to blockchain applications (e.g., \texttt{mutez} for tokens) as shown in Fig.~\ref{fig:types}, where \textbf{cpt} denotes comparable types. The type system does not support quantification over types and there are no user-defined types.
\begin{figure}[tp]
\begin{align*}
\textbf{cpt} &::= 
   \text{unit} 
   \Mid\text{never} 
   \Mid\text{bool} 
   \Mid\ \text{int}
   \Mid\text{nat}
   \Mid\text{string}
   \Mid\text{chain-id} 
    \Mid\text{bytes}
   \Mid\text{mutez} 
   \Mid\text{key-hash} \\
   &\Mid\text{key}
   \Mid \text{signature}
   \Mid \text{timestamp} 
   \Mid \text{address} 
   \Mid \text{option}\ \textbf{cpt}
   \Mid \text{or}\ \textbf{cpt}\ \textbf{cpt}
   \Mid \text{pair}\ \textbf{cpt}\ \textbf{cpt} \\
\text{T, U} &::= 
    \text{cpt}
   \Mid \text{option}\ \text{T}
   \Mid \text{list}\ \text{T}
   \Mid \text{set}\ \textbf{cpt} 
   \Mid \text{operation} 
   \Mid \text{contract}\ \text{T} \\
  & \Mid \text{pair}\ \text{T}\ \text{U}
   \Mid \text{or}\ \text{T}\ \text{U}
   \Mid \text{lambda}\ \text{T}\ \text{U} 
   \Mid \text{map}\ \textbf{cpt}\ \text{T}
   \Mid \text{big-map}\ \textbf{cpt}\ \text{T}
\end{align*}
\caption{Types}
\label{fig:types}
\end{figure}

Listing~\ref{lst:simple-program-add} contains a short example of a Michelson program. The code is interspersed with comments indicating the types of values currently on the stack. This program first duplicates its input (\lstinline|DUP|), decomposes the pair $P \times S$ into its components (\lstinline|UNPAIR|), compares the two values on top of the stack (\lstinline|COMPARE|), and transforms the result into a boolean (\lstinline|GE|). The \lstinline|IF| instruction pops the boolean and either executes the first (for true) or second (for false) list of instructions. At this point, only the initial pair remains on the stack. In the true branch, \lstinline|UNPAIR| decomposes the pair, and \lstinline|SUB| subtracts the two components. In the false branch, the \lstinline|PUSH| instruction puts a string message on the stack. The \lstinline|FAILWITH| instruction aborts the contract execution: the tokens provided with the invocation are consumed, but the storage and balance of the contract remain unchanged. Finally, the \lstinline|NIL| instruction pushes an empty list of operations, and \lstinline|PAIR| creates the required pair of type $O^* \times S$.
\lstset{language=michelson}
\begin{lstlisting}[float,captionpos=b,caption={A Michelson program},label={lst:simple-program-add}]
parameter int;
storage   int;
code {                (* pair int int :: [] *)
  DUP; UNPAIR;        (* int :: int :: pair int int :: [] *) 
  COMPARE; GE;        (* bool :: pair int int :: [] *)
  IF {UNPAIR; SUB} {PUSH string 'Unexpected Pair'; FAILWITH};
                      (* int :: [] *)
  NIL operation; PAIR (* pair (list operation) int :: [] *)}
\end{lstlisting}

A contract may have several entrypoints. Entrypoints are encoded by
using a sum type (called ``or type'' in Michelson). Each (nested)
alternative in a sum type can be
given a name, which is used as the name for an entrypoint. A call to
an entrypoint automatically adds the required \lstinline|LEFT| and
\lstinline|RIGHT| constructors of the sum type to the parameter.
\section{SCV: A Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}
SCV comes with a domain-specific language for specifying smart
contract properties, which is user-friendly for non-experts and compatible with SMT solvers. 
\subsection{Basic syntax}
\label{sec:basic-syntax}
Listing~\ref{lst:sub-contract-specification} shows the specification of the \texttt{sub} contract in the SCV language. Here, \KMCONTRACT, \KSPEC, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KINPUT{} and \KOUTPUT{} fields specify the input and output patterns, which are terms on Michelson values.

Figure~\ref{fig:syntax-terms-predicates} contains the syntax of terms: A term \textbf{t} can consist of constants of the sort \text{nat} (natural number), \text{int} (integer), \text{string} (string), and \text{b} (bytes) along with their respective types. The type annotation may be omitted except for the type \text{nat} to distinguish from integers. It also encompasses all blockchain constants \textbf{c}. When a variable appears for the first time, it must be accompanied by its type declaration. Subsequent mentions can omit the type declaration.

As Michelson operates on a stack, we assume that the input stack and the output stack each have exactly one element, as specified in the input term and the output term, respectively. The syntax allows us to specify either a fragment of Michelson code or a complete program. For a complete Michelson program, the output should be a pair consisting of an operation list and the state of the storage after code execution,  as described in Section \ref{sec:background}. For verification purposes, we elide the operation list and represent the output solely by the term of the value in the storage after execution concludes.

The pre- and postcondition instances are formulas in first-order logic that can be handled by an SMT solver (specifically, Z3 in our case). 
% The specification language is constructed from predicates. Formulas
% used in assertions or added to solvers are terms of Boolean
% type. Non-Boolean terms may also be used, and they can be mixed in
% any combination where types match appropriately. 
%
There are several predefined functions including \text{get\_map(e1, e2)}, which accesses an element in a map. If there is no key \text{e1} in the map \text{e2}, the function returns \text{None}; otherwise, it returns \text{Some(v)} where $v$ is the mapped value. Another function, \text{get\_contract(T, e)}, converts the address \text{e} to a contract on the Tezos blockchain. If no contract with an entrypoint of type \text{T} is associated with the address \text{e}, it returns \text{None}; otherwise, it returns the contract as \text{Some(c)}.

Certain functions are defined as predicates, such as \text{transfer\_token(e1, e2, e3)}. This predicate holds true if, after executing the smart contract code, there is an operation in the operation list indicating the transfer of \text{e2} tokens to address \text{e3} along with parameter \text{e1}\footnote{Additional predicates can be added if a mapping to Z3 formulas is defined.}.
%\lstset{language=michelson}
\begin{lstlisting}[float,captionpos=b,caption={Specification of  the \lstinline/sub/ contract},label={lst:sub-contract-specification},numbers=left]
mcontract sub = spec
  code := {DUP; UNPAIR; COMPARE; GE; IF {UNPAIR; SUB} 
          {PUSH string 'Unexpected Pair'; FAILWITH}}
  input := Pair (x : int) (y : int)
  output := (z : int)
  pre-condition := (x >= y)
  post-condition := (z = x - y) && (z >= 0)
\end{lstlisting}

In Listing \ref{lst:sub-contract-specification}, the program takes as input a stack with one

@@

The specification continues by defining the output term as a variable
\lstinline/z/ of type integer, a precondition stating
that the integer \lstinline/x/ is greater than or equal to
\lstinline/y/, and a postcondition asserting that the result

@@

Let us return to the auction contract. It is important to guarantee
that  the bid call can happen successfully after another if the new bid is higher than the previous
one (lines 16-17). The  \lstinline/%close/ entrypoint can be called
anytime after bidding if the caller is the contract owner (lines 18-19). However, it
should not be possible to call the  \lstinline/%bid/ entrypoint after
the auction is closed, and, of course, it cannot be closed
again if it is already closed (line 20).  Furthermore, contract deployment sets the  \lstinline/open/ flag to
\TRUE, indicating that the contract is open for bidding, i.e., users
should be able to bid on the contract or the contract owner should be
able to close it. Otherwise, the owner's original
deposit is lost. This sequence  illustrates
the life cycle of the auction contract, which is depicted in Figure
\ref{fig:auction-life-cycle}. Other properties related to the \lstinline|%close| entrypoint and the life cycle are omitted here due to space limitations.
\begin{figure}[tp]
    \centering
    \includegraphics[width=0.9\textwidth]{auction}
    \caption{Auction contract life cycle}
    \label{fig:auction-life-cycle}
\end{figure}

@@

A state is unreachable if its predicate \PREDICATE\ is  not
satisfiable. The following rule drops unreachable states.
\begin{mathpar}
\inferrule[]
  { \UNSAT\ \PREDICATE
  }{
  \{[\INSTRUCTION, \STACK, \PREDICATE]\} \cup \SYSTEM \SystemTrans \SYSTEM}
\end{mathpar}
The symbolic interpreter works on a graph with states as nodes and transitions as edges. This representation simplifies performing multi-step sub-transitions needed for some operations.

The semantics of the symbolic rules for the instructions adhere to the formal definition of the Michelson language as outlined in \cite{michelson,michelson1}. In our model, these instructions are classified based on how we handle them into one-step, multi-step, blockchain, cryptographic, branch, and loop instructions.

@@

Branch instructions, such as \IF\{\INSTRUCTIONONE\}\{\INSTRUCTIONTWO\} consume a value at the top of the stack. If the branch condition
is True, it executes \INSTRUCTIONONE; otherwise, it executes
\INSTRUCTIONTWO. 
\begin{mathpar}
  \inferrule[]
  { }{
    [(\IF\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TBOOL) \STACKCONCAT\STACK, \PREDICATE]
    \StateTrans\
    \{[\INSTRUCTIONONE; \INSTRUCTION, \STACK, \PREDICATE\ \Wedge\
    \StackOne],   [\INSTRUCTIONTWO; \INSTRUCTION, \STACK, \PREDICATE\ \Wedge\ \NEG\
   \StackOne]\}} 
\end{mathpar}

@@

We first look at the terminating loops, such as the \ITER\ instruction, which iterates over a
list. The typing rule for the \ITER\ instruction is as follows: 
\begin{mathpar}
  \inferrule{\JTypeExpr\TEnv{\INSTRUCTION}{\TY \STACKCONCAT \TYA\ \SRightarrow\ \TYA}
  }{
      \JTypeExpr\TEnv{\ITER\ \INSTRUCTION}{\TYLIST\ \TY \STACKCONCAT \TYA\ \SRightarrow\ \TYA}
    }
\end{mathpar}
This rule ensures that if the loop body \INSTRUCTION\ has the type \TY\ \STACKCONCAT\ \TYA\ \SRightarrow\ \TYA, then the entire \ITER\ instruction, when applied to a stack with the top element of type \TY\ \TYLIST\ and the rest of the stack of type \TYA, produces a result stack of type \TYA.

@@

In separate work \cite{dynamic-logic}, we have designed a dynamic logic tailored to Michelson as a formal foundation for the implementation of SCV. We have formalized a soundness proof of this logic in the Agda proof assistant.
\section{Static Checker}
\label{sec:static-checker-we}
The architecture of the static checker is illustrated in
Figure~\ref{fig:architecture-of-static-checker}. Users write
specifications in the SCV language, which are then parsed into an
Abstract Syntax Tree (AST) in OCaml. The input term along with the
code is passed to the symbolic interpreter, while the output term and
the pre- and postconditions are fed to the static checker. The
symbolic interpreter then runs the code with the provided input, where
the initial stack is derived from the input term.

@@

We have implemented approximately 85\% of the Michelson instructions,
including the most frequently used instructions. The remaining instructions were not implemented due to time constraints rather than technical limitations.

Consider the specification of the \lstinline/sub/ contract, the results of
the symbolic execution are either a failwith state with the path
condition \lstinline/x/ $<$ \lstinline/y/,  or a state in which the
final stack has only one element, which is a symbolic variable
\lstinline/sbvar/ of type integer, with the path condition stating
that the variable \lstinline/x/ is greater than or equal to
\lstinline/y/ and  \lstinline/sbvar/ is equal to
the subtract of \lstinline/x/ and \lstinline/y/. 
\subsection{Output check}
\label{sec:output-check}
For a fragment of Michelson code, the output term of the symbolic
execution has the form of a term \Term\ and its type \TY. For a
complete smart contract, it takes the form \PAIR\ \VOPERATIONLIST\
\VSTORAGE, \TPAIR\ (\TOPERATIONLIST) \TYS, where \VOPERATIONLIST\
represents the operation list and \VSTORAGE\ represents
the storage. Given
an output term specified as $\tau$, the checker scans all reachable
final states that do not contain a failwith stack and checks whether
$\tau$ matches the output term. In the former case, it matches
$\tau$ with the term \Term. In the latter case, it matches with
\VSTORAGE.

For instance, in the \lstinline|sub| contract, the output check ignores the states where the failwith occurs and verifies whether the output term \lstinline|z| matches the symbolic output term \lstinline/sbvar/ in the other state. It succeeds for this example as expected. However, this test could fail if, for example, the output term is incorrectly specified as a variable of a type that does not match the actual result.
\subsection{Fail conditions}
\label{sec:fail-conditions}
A Michelson contract can fail due to conditions introduced by the programmer, triggered by the \FAILWITH\ instruction along with an error message. Failures play a crucial role in smart contract programming to ensure that certain conditions lead to contract termination, such as unauthorized caller attempts (access control properties) or invalid arguments. Various smart contract standards, including \cite{erc,fa}, mandate specific fail conditions.

The static checker includes a feature to generate all fail conditions encountered in a smart contract. SCV identifies all instances where execution terminates due to the \FAILWITH\ instruction, collecting the associated path conditions and error messages. Users receive a comprehensive report detailing each fail condition along with its corresponding message. This capability enables verifying access control properties and other critical aspects across different smart contracts.
